---
title: "codici"
author: "Andrea Sciortino"
date: "2025-11-17"
output:
  pdf_document: default
  html_document: default
---

# Load libraries, data, functions and spatial graphs

```{r warning=FALSE, include=FALSE}
library(readr)
library(viridis)
library(tibble)
library(blackmarbler)
library(geodata)
library(sf)
library(terra)
library(ggplot2)
library(tidyterra)
library(lubridate)
library(readr)
library(dplyr)
library(spdep)
library(INLA)
library(gridExtra)
library(ggpubr)
library(igraph)
library(ggExtra)
library(patchwork)

enhanced_normalize_names <- function(x) {
  x <- tolower(iconv(x, to = "ASCII//TRANSLIT")) # remove accents, lower case
  x <- gsub("[.,]", "", x)                       # remove periods and commas
  x <- gsub("\\s*['`]\\s*", " ", x)              # normalize apostrophes and spacing
  x <- gsub("\\s+", " ", x)                      # normalize whitespace
  x <- gsub("^\\s+|\\s+$", "", x)                # remove leading/trailing whitespace
  x <- gsub("\\s*\\([^)]*\\)", "", x)            # remove content in parentheses
  x <- gsub("\\s*-\\s*", " ", x)                 # normalize hyphens
  x <- gsub("\\s+", "", x)                       # REMOVE ALL SPACES
  return(x)
}


compare_models <- function(fits, names, 
                          metrics = c("dic", "waic", "cpo", "mlik")) {
  
  results <- data.frame(Model = names)
  
  # Metric calculation functions
  metric_functions <- list(
    dic = function(fit) fit$dic$dic,
    waic = function(fit) fit$waic$waic,
    cpo = function(fit) -mean(log(fit$cpo$cpo), na.rm = TRUE),
    mlik = function(fit) fit$mlik[1, 1]
  )
  
  for(metric in metrics) {
    if(metric %in% names(metric_functions)) {
      results[[toupper(metric)]] <- sapply(fits, metric_functions[[metric]])
    }
  }
  return(results)
}

fit_inla_gamma <- function(formula, data) {
  # formula: R formula for the linear predictor
  # data: data.frame containing response and covariates 
  
  result <- inla(
    formula,
    family = "gamma",
    data = data,
    control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE),
    verbose = FALSE
  )
  
  return(result)
}
```

# Hierarchical Bayesian Model

In this section we describe the Bayesian hierarchical model three layes structure used to investigate latent field processes influencing the economic output and the relationship with nighttime lights.

-   ***Data Likelihood***: assumed the distribution Gamma, the two
    parameters $Y_i = Gamma( \alpha, b )$ define its distribution,
    where $\alpha$ is the shape parameter and $\beta$ is the rate
    parameter. We can the reformulate the distribution to obtain
    $E[Y_i] = \mu= \alpha / b$ and $Var[Y_i] = \alpha / b^2$
    $$ Y_i \sim Gamma(\mu, \sigma^2)$$

-   ***Model Process***: in order to incorporate information from the
    nighttime lights, the spacial structured effect and the unstructured
    effect, we used the logarithmic link function.

$$
log(E[Y_i]) = \eta_i = \beta_0 + \beta_1 \text{X}_{NTL} + \mathbf{v}_i
$$ 

where: (Fixed Effect)
- $\beta_0$ intercept
- $\beta_1$ linear parameter for regression associated to nighttime lights
- $X_i$ mean nighttime lights radiance for each province

(Structured Spatial Random Effect) 
- $\mathbf{v_i}$ has a ICAR distribution

$$\left[\mathbf{v_i} \mid \mathbf{v}_{j, j \neq i}, \tau_{v}^2\right] \sim N \left( \frac{1}{\sum_{j} w_{ij}} \sum_{j} v_j w_{ij},  \ \ \tau_{i}^2 = \frac{1}{\tau_{v}^2 \times \sum_{j} w_{ij}} \right)$$

-   ***Priors***: priors need to be choose, but assuming a
    non-informative prior will result in observing the data for
    inference without adding some prior belief on the distribution of
    the parameters.
    

### Data

```{r include=FALSE}
df.comuni = readRDS("~/Desktop/night-lights/comuni_data.rds")
df.province = readRDS("~/Desktop/night-lights/province_data.rds.rds")

df.comuni <- df.comuni %>%
  mutate(
    S = 1:n(),
    regione = as.numeric(as.factor(NAME_1))
  )

df.province <- df.province %>%
  mutate(
    year_idx = as.numeric(factor(year)), # 1-11 indexing
    space_time = 1:n(),             # Unique space-time IDs (for Type I)
    area_id_int = S,
    year_idx_int = year_idx,
    regione = as.numeric(as.factor(NAME_1))
  )

gdp.inla.graph = "~/gdp.graph.province"
gdp.inla.graph.comuni = "~/gdp.graph.comuni"
gdp.inla.graph.region = "~/gdp.graph.region"
gdp.inla.graph.province = "~/gdp.graph.PROV"

```

## Comuni

```{r}
model.com.1 = gdp ~ 1 + 
  ntl_mean 

model.com.2 = gdp ~ 1 + 
  ntl_mean +
  f(S, model="besag", graph=gdp.inla.graph.comuni)

fit1 = fit_inla_gamma(model.com.1, df.comuni)
fit2 = fit_inla_gamma(model.com.2, df.comuni)
```


```{r}

model1 = gdp ~ 1 + 
  ntl_mean +
  f(S, model="besag", graph=gdp.inla.graph.comuni) +
  f(regione, model="besag", graph = gdp.inla.graph.region, constr=TRUE)

fit3 <- inla(
    model1,
    family = "gamma",
    data = df.comuni,
    control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE),
    verbose = FALSE,
    control.predictor = list(compute = TRUE)
  )

summary(fit3)
```

```{r}
inla.priors.used(fit3)
```

## Province

```{r include=FALSE}
unique_provinces_sf <- df.province %>%
  group_by(name) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(name)  # Sort alphabetically by province name

# Create adjacency matrix
neighbors <- poly2nb(unique_provinces_sf)
adj_matrix <- nb2mat(neighbors, style = "B", zero.policy = TRUE)

# Use province NAMES (not S) as dimension names
province_names_sorted <- sort(unique(df.province$name))
rownames(adj_matrix) <- province_names_sorted
colnames(adj_matrix) <- province_names_sorted

# Create INLA graph
gdp.inla.graph1 <- inla.read.graph(adj_matrix)

df.province$name1 <- as.numeric(as.factor(df.province$name))
```

```{r}
model.province1 = gdp ~ 1 + ntl_mean

model.province2 = gdp ~ 1 + ntl_mean +
  f(name1, model="besag", graph = gdp.inla.graph1)

model.province3 = gdp ~ 1 + ntl_mean +
  f(name1, model="besag", graph = gdp.inla.graph1) +
  f(year, model = "rw1")

fit.pro1 = fit_inla_gamma(model.province1, df.province)
fit.pro2 = fit_inla_gamma(model.province2, df.province)
fit.pro3 = fit_inla_gamma(model.province3, df.province)
```

```{r}
summary(fit.pro3)
```


```{r}
model.provincea = gdp ~ 1 + ntl_mean + 
  f(year, model = "ar1")

model.provincerw = gdp ~ 1 + ntl_mean + 
  f(year, model = "rw1")

model.provinceaa = gdp ~ 1 + ntl_mean +
  f(name1, model="besag", graph = gdp.inla.graph1) +
  f(year, model = "ar1")
  
model.provinceaaa = gdp ~ 1 + ntl_mean +
  f(regione, model = "iid") +
  f(name1, model="besag", graph = gdp.inla.graph1) +
  f(year, model = "ar1")

fit.proa = fit_inla_gamma(model.provincea, df.province)
fit.proaa = fit_inla_gamma(model.provinceaa, df.province)
fit.proaaa = fit_inla_gamma(model.provinceaaa, df.province)
fit.provrw = fit_inla_gamma(model.provincerw, df.province)
```

```{r}
model_fits_province <- list(fit.pro1, fit.pro2, fit.pro3, fit.pro4, fit.proa, fit.proaa, fit.proaaa, fit.provrw)
model_names_province <- c("base", "spatial", "spatial + temp", "final", "temp AR1", "spatial + temp AR1", "spatial + region + temp AR1", "temp RW1")

comparison_province <- compare_models(model_fits_province, model_names_province )
comparison_province
```


```{r}
model.prov.final = gdp ~ 1 + ntl_mean +
  f(regione, model = "iid") +
  f(name1, model="besag", graph = gdp.inla.graph1) +
  f(year, model = "rw1")

fit.pro4 <- inla(
    model.prov.final,
    family = "gamma",
    data = df.province,
    control.compute = list(dic = TRUE, cpo = TRUE, waic = TRUE),
    verbose = FALSE 
  )

summary(fit.pro4)
```

```{r}
inla.priors.used(fit.pro4)
```

### Comparison of Models

```{r}
model_fits_comuni <- list(fit1, fit2, fit3)
model_names_comuni <- c("base", "spatial", "final")

comparison_comuni <- compare_models(model_fits_comuni, model_names_comuni)
comparison_comuni
```

```{r}
model_fits_province <- list(fit.pro1, fit.pro2, fit.pro3, fit.pro4)
model_names_province <- c("base", "spatial", "spatial + temp", "final")

comparison_province <- compare_models(model_fits_province, model_names_province, )
comparison_province
```

```{r}
# Models
# Normal-scale
model1 = gdp ~ 1 + f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale
model2 = log(gdp) ~ 1 +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale + night-light
model3 = log(gdp) ~ 1 + ntl_mean +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale + log(night-light)
model4 = log(gdp) ~ 1 + log(ntl_mean) +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale + log(night-light) + educations
model5 = log(gdp) ~ 1 + log(ntl_mean) + educations +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale + log(night-light) + employment
model6 = log(gdp) ~ 1 + log(ntl_mean) + employments +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale + log(night-light) + population
model7 = log(gdp) ~ 1 + log(ntl_mean) + populations +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)

# Logaritmic-scale + log(night-light) + population + education
model8 = log(gdp) ~ 1 + log(ntl_mean) + populations + employments +
  f(U, model='iid') + f(S, model="besag", graph=gdp.inla.graph)


fit1 = fit_inla_model(model1, df.prov, gdp.inla.graph)
fit2 = fit_inla_model(model2, df.prov, gdp.inla.graph)
fit3 = fit_inla_model(model3, df.prov, gdp.inla.graph)
fit4 = fit_inla_model(model4, df.prov, gdp.inla.graph)
fit5 = fit_inla_model(model5, df.prov, gdp.inla.graph)
fit6 = fit_inla_model(model6, df.prov, gdp.inla.graph)
fit7 = fit_inla_model(model7, df.prov, gdp.inla.graph)
fit8 = fit_inla_model(model8, df.prov, gdp.inla.graph)
```


# EDA

The core objective of this phase is to have a wide view of the data, use descriptive analysis to summaries it and choose a suitable distribution, witch will be fundamental for the further analysis. As the distribution of the variable represent an assumption we need to start from here.

### Economic Response Variable

```{r echo=FALSE}
# Histogram of GDP
p1 = ggplot(df.prov.2022, aes(x = gdp)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  labs(title = "Histogram of GDP", x = "GDP", y = "Count") +
  theme_minimal()

# Histogram of ntl_mean
p2 = ggplot(df.prov.2022, aes(x = log(gdp))) +
  geom_histogram(bins = 25, fill = "steelblue", color = "grey") +
  labs(title = "Histogram of Log(GDP)", x = "GDP", y = "Count") +
  theme_minimal()

p1 + plot_annotation(title = "Distribution of Economic Output: Province")
```

```{r echo=FALSE}
# Histogram of GDP
p1 = ggplot(df.comuni.2023, aes(x = gdp)) +
  geom_histogram(bins = 80, fill = "steelblue", color = "white") +
  labs(title = "Histogram of Total IRPEF Income", x = "Income", y = "Count") +
  theme_minimal()

p2 = ggplot(df.comuni.2023, aes(x = log(gdp))) +
  geom_histogram(bins = 50, fill = "blue2", color = "white") +
  labs(title = "Histogram of Log(Total IRPEF Income)", x = "Log(Income)", y = "Count") +
  theme_minimal()

p1 + p2 + plot_annotation(title = "Distribution of Economic Output: Municipality")
```

After this exploration we need to make assumption for distribution for the response variable, in order to fit the models, we decided as follow: 

 - provincial-level: due to the skewness of the data and issue
with symmetry, we will assume the Normal distribution for the initial
frequentist model for simplicity in order to avoid GLM, but in the
stochastic process model (Bayesian) the Gamma distribution better suit
for this case. 
 
 - municipality-level: for the same reason, the response
variable has been converted in logarithmic-scale, this procedure is
theoretically guaranteed and the distribution better suit for a Normal
distribution, there will be used for the model.

### Covariates

```{r echo=FALSE}
# Histogram of ntl_mean
p4 = ggplot(df.prov.2022, aes(x = log(ntl_mean))) +
  geom_histogram(bins = 35, fill = "darkgreen", color = "black") +
  labs(title = "Log(NTL Mean)", x = "Log(NTL Mean)", y = "Count") +
  theme_minimal()
# Histogram of ntl_mean
p3 = ggplot(df.prov.2022, aes(x = ntl_mean)) +
  geom_histogram(bins = 35, fill = "darkgreen", color = "black") +
  labs(title = "NTL Mean", x = "NTL Mean", y = "Count") +
  theme_minimal()

# Histogram of ntl_mean
p6 = ggplot(df.comuni.2023, aes(x = log(ntl_mean))) +
  geom_histogram(bins = 70, fill = "darkgreen", color = "black") +
  labs(title = "Log(NTL Mean) - Comuni", x = "Log(NTL Mean)", y = "Count") +
  theme_minimal()
# Histogram of ntl_mean
p5 = ggplot(df.comuni.2023, aes(x = ntl_mean)) +
  geom_histogram(bins = 70, fill = "darkgreen", color = "black") +
  labs(title = "NTL Mean - Comuni", x = "NTL Mean", y = "Count") +
  theme_minimal()

p3 + p4 + p5 + p6 + plot_annotation(title = "Distribution of the covariate: Province & Comuni")
```

# ADDITIONAL: Spatial Autocorrelation [ and addition frequentist traditional model ]
In spatial data analysis, it is common to encounter situations where we need to choose and construct the spatial structure in order to account for this conditional dependency.

## Neighborhood Structure

```{r}
nb.prov = poly2nb(df.prov.2022, snap = 15)  # Queen contiguity by default
nb2INLA("gdp.graph.prov", nb.prov)         
gdp.inla.graph = "gdp.graph.prov"

listw.prov = nb2listw(nb.prov, 
                      style = "W",  # Row-standardized
                      zero.policy = TRUE)  # Possible zero-length weights vectors

nb.comuni = poly2nb(df.comuni.2023, snap = 15)
nb2INLA("gdp.graph.comuni", nb.comuni)         
gdp.inla.graph.comuni = "gdp.graph.comuni"

listw.comuni = nb2listw(nb.comuni, 
                        style = "W", 
                      zero.policy = TRUE)
```

> The function builds a neighbours list based on regions with contiguous
> boundaries, that is sharing one or more boundary point.

## Spatial Dependency Measure

**Note**: We proceed to investigate spatial autocorrelation using
Moran's I statistic, it measure and quantifies the degree to which
similar values for a variable cluster together in space. The null
hypothesis test if values of the variable are distributed randomly
across the spatial units. Different assumption are made in order to
correctly interpreted this measure:

-   Stationary: the process governing the spatial dependence is assumed
    to be constant across the area
-   Weights dependency: the spatial autocorrelation do not exists in an
    absolute sense, it is only detectable under the specified
    neighborhood structure (adjacency).
-   Constant weight effect: equal spatial weights for one unit imply
    that all neighborhood equally influence the variable value
-   Residual: in order to demonstrate the violation of the linear
    regression assumption, we need to test the residual as assumed
    i.i.d. (different if we consider 'moran.test' or 'lm.morantest',
    first test the variable the other residuals)

> *Does the economic wealth/variable aggregated at province and
> municipality level shows a spatial autocorrelation ?*

```{r}
moran.test(
  df.prov.2022$gdp, 
  listw = listw.prov
  )

# moran.test(
#  df.comuni.2023$gdp, 
#  listw = listw.comuni
#  )

moran.test(
  log(df.comuni.2023$gdp), 
  listw = listw.comuni
  )
```

We verify the presence of autocorrelation in the response variable and spatially it do not show a random distribution across space. The positive coefficient is tested and statistical significant for both level of analysis.

```{r echo=FALSE}
# moran.plot(
#  df.comuni.2023$gdp, 
#  listw = listw.comuni
#  )

moran.plot(
  log(df.comuni.2023$gdp), 
  listw = listw.comuni
  )

moran.plot(
  df.prov.2022$gdp, 
  listw = listw.prov
  )
```

> The regression line is the global Moran’s Index. The spatially lagged
> value is a weighted average of the values from neighboring locations.

**Important**: Apparent autocorrelation, such as found in the previous
sections, may instead be caused by some underlying factors, here we
testing only the output variable. We can add predictors and complexity
to the model in order to answer: *What could be some
spatially-distributed causes of economic wealth ?*

Now we can check the linear model for spatial correlation of the
residuals, to test the violation of the linear model assumption.

```{r}
lm.morantest(
  lm_log_model.comuni,
  listw.comuni
)

lm.morantest(
  lm_model.prov,
  listw.prov
)
```

```{r eval=FALSE, include=FALSE}
# lm.morantest(
#  lm_model.comuni,
#   listw.comuni
# )
```

**Interpretation**: The probability that we would be wrong to reject the
null hypothesis of no spatial correlation in the residuals is extremely
low for the standard linear model at province-level and log-linear model
at municipality-level.

From this model on residual we are testing the indipendence of the residual, whitch show spatial correlation violating the indipendent assumption. 

## Autoregressive Models

### Simultaneous Autoregressive Models

In the linear model of the previous section we did not account for
spatial autocorrelation of the residuals, which indeed was present. So
the linear model violated one of its assumptions, i.e., independent
residuals. To account for this, we should refit the model as an
autoregressive model which accounts for spatial autocorrelation.

There are three main forms of Simultaneous Autoregressive (SAR) models,
with different explanations on the source of the autocorrelation of the
OLS residuals:

-   Spatial error : These imply that there are underlying
    spatially-correlated predictors which are not included in the linear
    model predictor list. Residuals are modeled by a regression on the
    residuals from adjacent areas. Formalized as $Y_i = X^T \beta + e_i$ 
    and the residual as $e_i = b(Y - X^T\beta) + \epsilon$, where
    $b$ the spatial dependency and $\epsilon$ represent the errors and
    are normal-distributed.

[code in another notebook, should I include ?]

-   Spatial lag : These imply that the response variable is influenced
    by the same response variable in neighboring areas. Formalized as
    $Y_i = X^T \beta + pWY + \epsilon_i$ where parameter $p$ controls
    the degree of autocorrelation of the response variable.

[code in another notebook, should I include ?]

-   Spatial Durbin: This model adds another spatial covariance
    parameter, $\nu$, which controls the degree of influence of the
    spatial lag of the covariates (predictors). Formalized as
    $Y_i = X^T \beta + pWY + WX\nu+ \epsilon_i$.

[Anselin Lesson Playlist](https://youtu.be/fdCNctqDyo4?si=9eNk2SNy3N1KQB48)

-   SAR (Simultaneous Autoregressive Models): Models a global,
    simultaneous spatial dependence process. The value at each location
    is assumed to be directly influenced by the values at all other
    locations in the system. Even if the $Y$ appear in the right-side of
    the equation, this form is not directly estimable via Ordinary Least
    Squares (OLS). Finding the solution for the coefficients involves
    solving a linear system that accounts for the spatial structure of
    the residuals. It is called simultaneous because the values across
    all locations are determined jointly.

-   CAR (Conditional Autoregressive Model): Models a local, conditional
    spatial dependence. It specifies that the value at a location, given
    the values of its immediate neighbors, is independent of all other
    locations. This is a **Markov Random Field**, and so the Bayesian
    framework better apply in this context.


