---
title: "Nighttime Lights"
author: "Andrea Sciortino"
date: "2025-09-13"
output: html_document
---

# Spatial Statistics Data Preparation for Nighttime Lights in Italy

## Areal Data Preparation

Need to define the geographical domain with proper object and the we can retrieve both the raster of the total nighttime light emission caputred across the surface. We need the mean value across all the pixel enclosed by the administrative boundary of the region of interest.

```{r}
library(ggplot2)
library(tidyterra)
library(tibble)
library(blackmarbler)
library(geodata)
library(sf)
library(terra)
library(readr)
library(dplyr)
source("~/Desktop/night-lights/tokens.R")

normalize_names <- function(x) {
  x <- tolower(iconv(x, to = "ASCII//TRANSLIT")) # remove accents, lower case
  x <- gsub("\\s+", " ", x)      # normalize whitespace
  x <- trimws(x)                 # remove leading/trailing whitespace
  return(x)
}
```

# Province Data

Define the geographical domain with a `sf` object, then the location for the areal data

```{r}
# DATA PREPARATION - Nighttime Lights Italy
# Technical: NASA Black Marble VIIRS VNP46A4 annual product
# Spatial resolution: ~500m, Temporal coverage: 2012-present

# 1. ADMINISTRATIVE BOUNDARIES
italia <- gadm("ITA", 1, path=tempdir())        # Regions (NUTS2)
provinces <- gadm("ITA", 2, path=tempdir())     # Provinces (NUTS3)
comuni <- gadm("ITA", 3, path=tempdir())        # Municipalities (LAU)

# 2. NIGHTTIME LIGHTS EXTRACTION
# Annual mean radiance (nW/cm²/sr) extraction using blackmarbler
# Note: Requires NASA Earthdata bearer token
ntl_province = bm_extract(
  roi_sf = provinces,
  product_id = "VNP46A4",      # Annual VIIRS product
  date = 2012:2022,            # Time series
  bearer = bearer,             
  stats = "mean"               # Aggregation statistic
)
```

### Trends over time

Analyses of time series data aggregated over spatial regions (areal data) need to model both temporal dynamics and spatial dependence, as ignoring spatial context can lead to biased inference and imprecise predictions

```{r}
str(ntl_df)
summary(ntl_df)
# table(ntl_df$NAME_1)

ggplot(ntl_df, aes(x = date, y = ntl_mean, color = NAME_2)) +
  geom_line() + 
  labs(title = "Annual Mean Nighttime Lights by Province and Region",
       x = "Year", y = "NTL Mean") +
  theme_minimal() +
  
  theme(legend.position = "none")
```


## Visualization

```{r}
#### Make raster
raster = bm_raster(roi_sf = italia,
               product_id = "VNP46A4",
               date = 2023,
               bearer = bearer)

r = raster |> terra::mask(italia)

## Distribution is skewed, so log
r[] = log(r[] + 1)

##### Map
ggplot() +
  geom_spatraster(data = r) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  legend.position = "none")
```

```{r}
ntl_2022 <- bm_extract(
  roi_sf = provinces,
  product_id = "VNP46A4",   # Annual data
  date = 2022,              # Year of interest
  bearer = bearer           # Your token
)

ntl_2023.comuni <- bm_extract(
  roi_sf = comuni,
  product_id = "VNP46A4",   # Annual data
  date = 2023,              # Year of interest
  bearer = bearer,          # Your token
  quality_flag_rm = 1
)

str(ntl_2023.comuni)
summary(ntl_2023.comuni)

str(ntl_2012)
summary(ntl_2012)
```

Visualization as areal data.

```{r}
library(viridis)

ntl.prov.new = ntl_2022 |> st_as_sf()

# Create the visualization
ggplot(data = ntl.prov.new) +
  geom_sf(aes(fill = ntl_mean)) + # Draw regions with fill color based on NTL value, no border
  scale_fill_viridis(
    name = "Radiance", # Legend title
    option = "D",
  ) +
  theme_minimal() # Clean, minimal theme

```



```{r}

library(ggplot2)
library(patchwork)

ggplot(ntl_2022, aes(x = ntl_mean)) +
  geom_histogram(binwidth = 0.9, fill = "orange", color = "white") +
  labs(title = "Distribution of Mean NTL (2022)", x = "Mean Nighttime Light", y = "Count") +
  theme_minimal()

# Visualization ggregated for region:
# We can see the region with the most values of radiance

ggplot(ntl_2022, aes(x = reorder(NAME_1, ntl_mean), y = ntl_mean)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Regional NTL Variation (2022)", x = "Region", y = "Mean Nighttime Light") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This is the complete data preparation for the provincial level nighttime lights data for Italy. Next step is to join with the spatial object for the provinces, linked to the responce variable.

# Comuni Data

## Data Preparation

```{r eval=FALSE, include=FALSE}
# Comuni shapefile from ISTAT official website
comuni = st_read("Com01012025/Com01012025_WGS84.shp")

ntl_comuni <- bm_extract(
  roi_sf = comuni,                  # sf with province polygons
  product_id = "VNP46A4",           # annual product
  date = 2022,                      # year of interest
  bearer = bearer,                  # NASA API token
  aggregation_fun = "mean" )         # statistic to calculate per province

ntl_comuni |> left_join(comuni) |> st_as_sf()

str(comuni)

# Visualizzation
ggplot(ntl_comuni) +
  geom_sf(aes(fill = ntl_mean), color = NA) +
  scale_fill_viridis_c(option = "E", name = 'Radiance') +
  theme_void() +
  labs(fill = "Night-time Intensity")
```





## [ Temporal Extraction ] not involved in the analysis but useful for future reference
```{r}
date_seq <- seq.Date(from = as.Date("2018-01-01"), to = as.Date("2018-04-01"), by = "month")

ntl_df.prova <- bm_extract(
  roi_sf = provinces,
  product_id = "VNP46A3",
  date = date_seq,      # sequence of monthly dates
  bearer = bearer
)

ggplot(ntl_df.prova, aes(x = date, y = ntl_mean, color = NAME_2)) +
  geom_line() + 
  labs(title = "Annual Mean Nighttime Lights by Province and Region",
       x = "Year", y = "NTL Mean") +
  theme_minimal() +
  theme(legend.position = "none")
```

-   Sparse Matrix: Most points are likely empty (white) because provinces have relatively few direct neighbors; this is typical for geographic adjacency.